---
title: "Mycobiome Exploration"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Introduction

### Rationale for exploratory association analyses

How to integrate microbiome, dietary, and clinical metadata to identify microbial signatures associated with inflammation or dietary patterns.

Continuous-by-continuous comparisons?

-   Scatterplots to examine whether fungal diversity or candidate fungal taxa vary with continuous dietary or character variables.

-   diversity x fiber intake, candidate fungal abundance x nutrient variables

-   Inflammation / IBD potential candidates: increase in *Candida albicans*, decrease in *Saccharomyces cerevisiae*, *Saccharomyces*, *Candida*, and the Basidiomycota:Ascomycota ratio.

Categorical-by-continuous comparisons?

-   Box plots of fungal diversity or candidate taxa across disease activity groups or HBI score?

Candidate fungal features in literature: *Candida albicans*, *Saccharomyces cerevisiae*, *Saccharomyces*, *Candida*, and the Basidiomycota:Ascomycota ratio.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(plotly)
library(here)
library(readr)
library(readxl)

# ── Data sources ─────────────────────────────────────────────
taxa_long  <- readRDS(here("data", "intermediate", "taxa_long_list.rds"))
alpha_long <- readRDS(here("data", "intermediate", "alpha_long.rds"))
merged     <- read_csv(here("data", "processed", "merged.csv"),
                       show_col_types = FALSE)

# ── Column groups ────────────────────────────────────────────
meta_cols <- c("Sample_ID", "Participant_ID", "Sample_type",
               "Study_group_new", "Fiber_restriction")

# Dietary exposures of interest
diet_vars <- c(
  "Fiber (g)"               = "TotFib (g)",
  "Tryptophan (g)"          = "Trp (g)",
  "Artificial sweeteners (mg)" = "ArtSw (mg)",
  "Calories (kcal)"         = "Cals (kcal)"
)

# Continuous participant characteristics
char_vars <- c(
  "Age"                  = "age",
  "BMI"                  = "bmi_1",
  "Harvey-Bradshaw Index" = "harvey_bradshaw_index"
)

# All x-axis options for scatter
x_vars <- c(diet_vars, char_vars)

# Mycobiome candidates for y-axis
myco_candidates <- c(
  #Commensalism and pathogenic, protective of immune function during
  #times of stress; associated IBD paitents and inflammation
  "C. albicans (species)"         = "s__albicans",
  # Surrogate marker of inflammation: lower in non-inflammed areas
  "S. cerevisiae (species)"       = "s__cerevisiae",
  "Saccharomyces (genus)"         = "g__Saccharomyces",
  # Higher in IBD flare (Gibberella, monoliformis,
  # Alternaria brassicicola, cryptococcus neoformans)
  "Candida (genus)"               = "g__Candida",
  # Ratio increase in IBD compared to control
  "Basidio:Asco ratio"            = "ratio_basidio_asco"
)

# Disease status colour palette
group_pal <- c(
  "Active IBD"  = "#D62728",
  "Quiescent"   = "#FF7F0E",
  "Non-IBD"     = "#2CA02C"
)

# HBI activity category palette
hbi_pal <- c(
  "Remission"       = "#2CA02C",
  "Mild"            = "#FF7F0E",
  "Moderate/Severe" = "#D62728"
)

# ── Derived columns─────────────────────────────────────────────
merged <- merged |>
  mutate(
    ratio_basidio_asco = p__Basidiomycota / (p__Ascomycota + 1e-6)
  )

```

# Checking participants

```{r}

check <- distinct(merged, Sample_ID, Participant_ID, Study_group_new)
check <- distinct(alpha_long, Sample_ID, Participant_ID, Study_group_new)

```

------------------------------------------------------------------------

# Population

## Scatter — dietary / characteristics × diversity or mycobiome

Potential indicators of inflamation?

```{r scatter-setup}
# Join merged (wide, per sample) with alpha diversity
alpha_wide <- alpha_long |>
  select(Sample_ID, Diversity_metric, Value) |>
  pivot_wider(names_from = Diversity_metric, values_from = Value)

scatter_df <- merged |>
  left_join(alpha_wide, by = "Sample_ID") |>
  rename(Disease_status = Study_group_new)

# Rename raw cols to display names for selectors
scatter_df <- scatter_df |>
  rename(
    `Fiber (g)`                  = `TotFib (g)`,
    `Tryptophan (g)`             = `Trp (g)`,
    `Artificial sweeteners (mg)` = `ArtSw (mg)`,
    `Calories (kcal)`            = `Cals (kcal)`,
    Age                          = age,
    BMI                          = bmi_1,
    `Harvey-Bradshaw Index`      = harvey_bradshaw_index
  ) |>
  mutate(
    HBI_category = case_when(
      `Harvey-Bradshaw Index` <= 4 ~ "Remission",
      `Harvey-Bradshaw Index` <= 7 ~ "Mild",
      `Harvey-Bradshaw Index` > 7 ~ "Moderate/Severe"
    ),
    HBI_category = factor(HBI_category,
                          levels = c("Remission", "Mild", "Moderate/Severe"))
  )

check <- distinct(merged, Sample_ID, Participant_ID, Study_group_new)

```

```{r scatter-fn}
# Helper: one scatter + linear fit coloured by disease status
plot_scatter <- function(df, x_col, y_col, x_label = x_col, y_label = y_col) {
  # Filter
  plot_data <- df |> filter(!is.na(.data[[x_col]]), !is.na(.data[[y_col]]))
 
  # Plot
  p <- plot_data |>
    ggplot(aes(
      x      = .data[[x_col]],
      y      = .data[[y_col]],
      colour = Disease_status,
      text   = paste0(
        "ID: ", Participant_ID, "<br>",
        x_label, ": ", round(.data[[x_col]], 2), "<br>",
        y_label, ": ", round(.data[[y_col]], 4)
      )
    )) +
    geom_point(size = 2.5) +
    geom_smooth(method = "lm", se = TRUE, aes(group = 1),
                colour = "grey40", linewidth = 0.7) +
    scale_colour_manual(values = group_pal, name = "Disease status") +
    labs(x = x_label, y = y_label) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
}
```

### Fiber × diversity

```{r scatter-fiber-diversity}
plot_scatter(scatter_df, "Fiber (g)", "Shannon",
             y_label = "Shannon diversity")

# 2 NA's
check <- distinct(scatter_df, Disease_status, `Fiber (g)`, Sample_ID)
```

### Tryptophan × diversity

```{r scatter-trp-diversity}
plot_scatter(scatter_df, "Tryptophan (g)", "Shannon",
             y_label = "Shannon diversity")
```

### Artificial sweeteners × diversity

```{r}
unique(scatter_df$`Artificial sweeteners (mg)`)
unique(scatter_df$ratio_basidio_asco)
```

```{r scatter-artsw-albicans}
# All NA's
# plot_scatter(scatter_df, "Artificial sweeteners (mg)", "Shannon",
#              y_label = "Shannon")
```

### Age × Diversity

```{r scatter-age-diversity}
plot_scatter(scatter_df, "Age", "Shannon",
              y_label = "Shannon")
```

------------------------------------------------------------------------

## Boxplots — study group distributions

```{r box-fn}
plot_box <- function(df, y_col, y_label = y_col) {
  plot_df <- df |> filter(!is.na(.data[[y_col]]))

  p <- plot_df |>
    ggplot(aes(
      x    = Disease_status,
      y    = .data[[y_col]],
      fill = Disease_status,
      text = paste0("ID: ", Participant_ID, "<br>",
                    y_label, ": ", round(.data[[y_col]], 4))
    )) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.08, size = 2, alpha = 0.8) +
    scale_fill_manual(values = group_pal, guide = "none") +
    labs(x = NULL, y = y_label) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
}
```

```{r}
table(scatter_df$Disease_status, scatter_df$`Harvey-Bradshaw Index`)
```

```{r}
table(scatter_df$Disease_status)
```

### Shannon diversity by study group

```{r box-shannon}
plot_box(scatter_df, "Shannon", "Shannon diversity")
```

### C. albicans by study group

```{r box-albicans}
plot_box(scatter_df, "s__albicans", "C. albicans abundance")
```

### S. cerevisiae by study group

```{r box-scer}
plot_box(scatter_df, "s__cerevisiae", "S. cerevisiae abundance")
```

### Basidio:Asco ratio by study group

```{r box-ratio}
plot_box(scatter_df, "ratio_basidio_asco", "Basidiomycota / Ascomycota")
```

### Fiber by study group

```{r box-fiber}
plot_box(scatter_df, "Fiber (g)", "Total fiber (g)")
```

------------------------------------------------------------------------

## Boxplots — HBI activity category

### HBI distribution by disease status

```{r hbi-ridge, message=FALSE, warning=FALSE}
scatter_df |>
  # mutate(`Harvey-Bradshaw Index` =
  #          replace_na(`Harvey-Bradshaw Index`, 0)) |>
  filter(Disease_status != "Non-IBD") |> 
  ggplot(aes(
    x    = `Harvey-Bradshaw Index`,
    y    = Disease_status,
    fill = Disease_status
  )) +
  geom_density_ridges(alpha = 0.6) +
  scale_fill_manual(values = group_pal, guide = "none") +
  scale_x_continuous(limits = c(0, 10)) +
  labs(x = "Harvey-Bradshaw Index", y = NULL) +
  theme_minimal()
```

```{r hbi-box-fn}
plot_hbi_box <- function(df, y_col, y_label = y_col) {
  plot_data <- df |>
    filter(Disease_status != "Non-IBD",
           !is.na(HBI_category),
           !is.na(.data[[y_col]]))
 
  p <- plot_data |>
    ggplot(aes(
      x    = HBI_category,
      y    = .data[[y_col]],
      fill = HBI_category,
      text = paste0(
        "ID: ", Participant_ID, "<br>",
        "HBI: ", `Harvey-Bradshaw Index`, " (", HBI_category, ")<br>",
        y_label, ": ", round(.data[[y_col]], 4)
      )
    )) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.08, size = 2, alpha = 0.8) +
    scale_fill_manual(values = hbi_pal, guide = "none") +
    labs(x = "HBI activity category", y = y_label) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
}
```

### Basidio:Asco ratio by HBI category

```{r hbi-box-ratio}
plot_hbi_box(scatter_df, "ratio_basidio_asco", "Basidiomycota / Ascomycota")
```

### C. albicans by HBI category

```{r hbi-box-albicans}
plot_hbi_box(scatter_df, "s__albicans", "C. albicans abundance")
```

### S. cerevisiae by HBI category

```{r hbi-box-scer}
plot_hbi_box(scatter_df, "s__cerevisiae", "S. cerevisiae abundance")
```

### Shannon diversity by HBI category

```{r hbi-box-shannon}
plot_hbi_box(scatter_df, "Shannon", "Shannon diversity")
```

------------------------------------------------------------------------

# Individual

## Mycobiome composition per patient
